Katana: An ELF/DWARF Manipulation Tool with Hotpatching Capabilities

### 2011-04-18 doc/katana.tex
updated documentation for 0.2 release

% Created 2011-04-13 Wed 00:01
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{graphicx}

\title{Katana: An ELF/DWARF Manipulation Tool with Hotpatching Capabilities}
\author{James Oakley}
\date{13 April 2011}

@@

    A COMMAND is of the form COMMAND$_{\mathrm{IDENTIFIER}}$ PARAM PARAM PARAM \ldots{}., where
    tokens are seperated by spaces and the number of PARAMs depends on
    the command.

@@

    A COMMAND$_{\mathrm{IDENTIFIER}}$ is one or more words which identify a
    COMMAND. In many cases a command is identified by only one word,
    but sometimes similar commands are grouped by sharing the first
    word in their identifier.

@@

     The following types of variables exist
\begin{itemize}
\item string
\item elf
\item elf section
\item raw data
\item array
\end{itemize}
\end{itemize} % ends low level
\subsubsection{Available Commands}
\label{sec-2.1.2}

@@

      \emph{Usage}: \texttt{dwarfscript emit [SECTION] ELF OUTFILE}\\
      \emph{Params}: SECTION must be the name (string) of the section to write as
                Dwarfscript. If not specified it defaults to
                ``.eh$_{\mathrm{frame}}$''. ELF must be an ELF object. OUTFILE must
                be a string with the name of a file to write the resulting
                Dwarfscript to.\\
      \emph{Function}: Writes the Dwarfscript representation of the given
                  SECTION from the given ELF to OUTFILE.

@@

      \emph{Usage}: \texttt{dwarfscript compile INFILE}\\
      \emph{Params}: INFILE must be a string containing the name of a file.\\
      \emph{Function}: Interprets the contents of the file named by INFILE
                  as Dwarfscript and compiles the Dwarfscript into
                  beinary form. Returns an array with 3 items
                  0: raw data for .eh$_{\mathrm{frame}}$ 
                  1: raw data for .eh$_{\mathrm{frame}}$$_{\mathrm{hdr}}$
                  2: raw data for .gcc$_{\mathrm{except}}$$_{\mathrm{table}}$.
\end{itemize} % ends low level

@@

      \emph{Usage}: \texttt{replace section ELF SECTION\_NAME NEW\_SECTION}
      \emph{Params}: ELF must be an ELF object. SECTION$_{\mathrm{NAME}}$ must be a
                string. NEW$_{\mathrm{SECTION}}$ must be either an ELF section or raw data.
      \emph{Function}: Replaces the section with the name SECTION$_{\mathrm{NAME}}$ in
                  the oject ELF with the data from
                  NEW$_{\mathrm{SECTION}}$. Section headers are replaced if NEW$_{\mathrm{SECTION}}$ is
                  able to provide them, but not if it is only raw data.

@@

      \emph{Usage}: \texttt{replace raw ELF OFFSET NEW\_DATA}
      \emph{Params}: ELF must be an ELF object. ADDRESS must be an
                integer. NEW$_{\mathrm{DATA}}$ must be raw data.
      \emph{Function}: Replaces the raw data at OFFSET in the ELF object
                  with NEW$_{\mathrm{DATA}}$. OFFSET must refer to a location in an
                  existing section.
\end{itemize} % ends low level

@@

      \emph{Usage}: \texttt{hash elf STR}
      \emph{Params}: STR must be a string.
      \emph{Function}: Prints the result of running elf$_{\mathrm{hash}}$ (from libelf)
                  on the string.

@@

@@

    the katana manpage (although the information in this document is
    considerably more extensive than in the manpage)
\subsection{Patch Object Format}
\label{sec-3.6}

   This section of the document is not yet written. It will provide a description and specification of the PO format used by Katana
\subsection{Patch Generation Process}
\label{sec-3.7}

   This section of the document is still under construction. When
   complete, it will provide a description of the internal process that
   Katana uses to generate a patch. Understanding it is not necessary
   for using Katana.

@@

@@

\begin{itemize}
\item Katana compare the old and new source trees, looking at the object (.o)
     files.
\item For object files which exist only in the new tree, their contents
     are added to the patch object being created.
\item For object files which exist only in the old tree, a warning
     about their removal is issued and nothing further is done.
\item For object files which exist in both trees, type diffing and
     function diffing are performed and the differences are written
     tot he patch object being created.
\end{itemize}
\subsection{Type Diffing}
\label{sec-3.11}

   This section of the document still needs to be written. The general
   idea is that structures are examined for for added members, moved
   members, and changed members.
\subsection{Function Diffing}
\label{sec-3.12}

\subsection{Patch Application Process}
\label{sec-3.13}

   This section of the document is not yet written. It will provide a
   description of the internal process that Katana uses to apply a
   patch. Understanding it is not necessary for using Katana.
\subsection{Roadmap}
\label{sec-3.14}

@@

\section{Credits and Licensing}
\label{sec-5}

@@

  Katana is being written by James Oakley and was designed by Sergey
  Bratus, Ashwin Ramaswamy, James Oakley, Michael Locasto, and Sean
  Smith.

>>>>

% Created 2011-04-18 Mon 01:44
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{graphicx}

\title{Katana: An ELF/DWARF Manipulation Tool with Hotpatching Capabilities}
\author{James Oakley}
\date{18 April 2011}

@@

    A COMMAND is of the form COMMAND\_{}IDENTIFIER PARAM PARAM PARAM \ldots{}., where
    tokens are seperated by spaces and the number of PARAMs depends on
    the command.

@@

    A COMMAND\_{}IDENTIFIER is one or more words which identify a
    COMMAND. In many cases a command is identified by only one word,
    but sometimes similar commands are grouped by sharing the first
    word in their identifier.

@@

     The following types of variables exist
\begin{itemize}
\item string
\item ELF
\item ELF section
\item raw data
\item array
\end{itemize}
\end{itemize} % ends low level
\subsubsection{Available Commands}
\label{sec-2.1.2}

@@

      \emph{Usage}: \texttt{dwarfscript emit [SECTION] ELF OUTFILE}\\
      \emph{Params}: SECTION must be the name (string) of the section to write as
                Dwarfscript. If not specified it defaults to
                ``.eh\_{}frame''. ELF must be an ELF object. OUTFILE must
                be a string with the name of a file to write the resulting
                Dwarfscript to.\\
      \emph{Function}: Writes the Dwarfscript representation of the given
                  SECTION from the given ELF to OUTFILE.

@@

      \emph{Usage}: \texttt{dwarfscript compile INFILE}\\
      \emph{Params}: INFILE must be a string containing the name of a file.\\
      \emph{Function}: Interprets the contents of the file named by INFILE
                  as Dwarfscript and compiles the Dwarfscript into
                  beinary form. Returns an array with 3 items
                  0: raw data for .eh\_{}frame 
                  1: raw data for .eh\_{}frame\_{}hdr
                  2: raw data for .gcc\_{}except\_{}table.
\end{itemize} % ends low level

@@

      \emph{Usage}: \texttt{replace section ELF SECTION\_NAME NEW\_SECTION}
      \emph{Params}: ELF must be an ELF object. SECTION\_{}NAME must be a
                string. NEW\_{}SECTION must be either an ELF section or raw data.
      \emph{Function}: Replaces the section with the name SECTION\_{}NAME in
                  the oject ELF with the data from
                  NEW\_{}SECTION. Section headers are replaced if NEW\_{}SECTION is
                  able to provide them, but not if it is only raw data.

@@

      \emph{Usage}: \texttt{replace raw ELF OFFSET NEW\_DATA}
      \emph{Params}: ELF must be an ELF object. ADDRESS must be an
                integer. NEW\_{}DATA must be raw data.
      \emph{Function}: Replaces the raw data at OFFSET in the ELF object
                  with NEW\_{}DATA. OFFSET must refer to a location in an
                  existing section.
\end{itemize} % ends low level

@@

      \emph{Usage}: \texttt{hash elf STR}
      \emph{Params}: STR must be a string.
      \emph{Function}: Prints the result of running elf\_{}hash (from libelf)
                  on the string.

@@

    Note that this feature is a work in progress. There isn't much you
   can do with configuration files right now and the information here
   may be out of date. Please do not rely on it.

@@

\begin{itemize}
\item The katana manpage (although the information in this document is
      considerably more extensive than in the manpage)
\item S. Bratus, J. Oakley, A. Ramaswamy, S. Smith,
      M. Locasto. \emph{Katana: Towards Patching as a Runtime part of the       Compiler-Linker-Loader Toolchain}. International Journal of
      Secure Software Engineering (IJSEE). 1, 3 (2010).
\end{itemize}
\subsection{Patch Object Format}
\label{sec-3.6}

   We have developed a patch object (PO) format which we hope will
   eventually pave the way for a standardized vendor-neutral patch
   format for hotpatching. We are not advancing our format as such,
   but it embodies some of the principles which we think are
   important. Why should patching not be a part of the ABI and of the
   standard toolchain?
\begin{itemize}
\item A PO is a valid ELF file.
\item A PO utilizes DWARF information to describe types, variables, and
     functions requiring patching.
\item A PO allows type transformations to be specified using a language
     based on the DWARF standard.
\end{itemize}
   Through the use of existing standards and well-structured ELF files
   utilizing a simple expression language for data patching, we aim to
   create patches that are easily examined (or modified) with existing
   tools. Relocatable objects containing new code and data which may
   be inserted at runtime are nothing new. This is the entire premise
   of the dynamic library. User-written functions which may have this
   code injection (in the case of patching data where the desired
   actions cannot be determined automatically) already exist as the
   .init and .fini sections. It is our view, however, that it is
   important to have a seperate patch format as opposed to patches
   merely being dynamic libraries which contain both the patch data
   and the logic to perform the patching (as is done by some other
   hotpatching systems). We view this as an unnecessary mixing of data
   and logic. The code to apply patches should live in one place on
   any given system, as most other executable content does.

   As an ELF object, our PO files contain the following non-standard
   sections.

\begin{itemize}
\item .text.new
     Contains new/modified functions
\item .rodata.new, .data.new
     new data
\item .unsafe\_{}functions
     Contains a simple listing (of symbol indices) of the functions in
     the binary to be patched which should not have activation records
     on the stack when patching is taking place.
\item .debug\_{}info
     Contains listings of the variables and functions which need to be
     patched using the DWARF data format. This section is standard and
     is used here with validly formatted data, but is used for
     patching instead of debugging. The use of the the .debug\_{} name is
     preserved for compatibility with libdwarf and tools such as
     readelf, objdump, dwarfdump capable of listing DWARF
     information. It can be, however, confusing and the name will
     likely change in the future.
\item .debug\_{}frame
     Like .debug\_{}info a standard section used in a nonstandard way,
     see notes above about the naming. Contains an extended version of
     DWARF Call Frame Information which describes how various data
     structures are to be patched. The details are not properly
     documented at the moment, please email the Authors for more
     details if you would like further information.
\end{itemize}
\subsection{Patch Generation Process}
\label{sec-3.7}

   This section of the document is still under construction, but we
   hope that the information that is provided will be of some use.

@@

   Note that this feature is a work in progress. There isn't much you
   can do with configuration files right now.

@@

   High level view:
\begin{itemize}
\item Katana compare the old and new source trees, looking at the object (.o)
     files.
\item For object files which exist only in the new tree, their contents
     are added to the patch object being created.
\item For object files which exist only in the old tree, a warning
     about their removal is issued and nothing further is done.
\item For object files which exist in both trees, type diffing and
     function diffing are performed and the differences are written
     tot he patch object being created.
\end{itemize}
   A more detailed (although still very rough) algorithm:

\begin{verbatim}
Walk the old and new object trees in parallel
  For each pair of objects (corresponding old and new objects)
    If the new object does not exist
      Issue a warning and continue
    If the old object does not exist
      Add all functions and vars to patch
      Continue
    If the two objects are the same
      Continue
    If the two objects differ
      For every global variable in the old object
        Compare with matching variable in the new object
        If the two are a different type or the type struct changes
          Generate a type transformation for the patch
        If the variable initializers changed
          If the variable is const
            Add new data to the patch
          Else
            Generate a warning (can't determine automatically if
                the change should be applied)
        If anything related to the variable changed
          Find all functions using the variable
          Add them to the unsafe functions list
      For every global variable only in the new object
        Add it to the patch
      For every function in the old object
        Compare with matching function in the new object
          If the functions differ
            Add the new text to the patch
            Add the function to the unsafe functions list.
      For ever function only in the new object
        Add the function to the patch
Write out the patch ELF!
\end{verbatim}

\subsection{Type Diffing}
\label{sec-3.11}

   The general idea is that structures are examined for for added
   members, moved members, and changed members. If you need more
   detail than this, please contact the Authors.
\subsection{Function Diffing}
\label{sec-3.12}

   Functions are compared in an unsophisticated manner. The comparison
   is essentially byte-by-byte (i.e. no parsing of the machine
   instruction set is done). If bytes differ between the compiled
   version of the old function and the new function, then the function
   is assumed to need patching.  The one exception to this is that
   relocations are accounted for. If bytes differ at an address that
   is fixed up by relocations, the relocations are examined to make
   sure that they are for the same symbol. If in fact they are, then
   the function is deemed not to have changed. If the symbol referred
   to corresponds to a variable that has changed then it may need to
   be moved to be patched. In that event the function may in fact have
   to be modified, but it will be modified only to apply the
   relocations rather than as a patch to the function per se and thus
   the function diffing stage does not concern itself with whether
   referenced symbols have changed.
\subsection{Patch Application Process}
\label{sec-3.13}

   This section of the document is not yet written. It will provide a
   description of the internal process that Katana uses to apply a
   patch. Understanding it is not necessary for using Katana.

   The basic process is as follows

\begin{verbatim}
Read the patch file
Calculate versioning. This is currently not implemented.
Find malloc in the target, as we may need it
Calculate a safe state for the target (based on the unsafe functions list)
Wait for the target to reach a safe state
Map in necessary sections from the patch
Copy PLT and GOT to new locations as we may need to expand them
For each variable listed in the patch
  Apply the variable patch
For each function listed in the patch
 Apply the function patch
Apply necessary relocations
\end{verbatim}

\subsection{Roadmap}
\label{sec-3.14}

@@

  For information on how katana can be used specifically for DWARF
  manipulation, please see Dartmouth College Tech Report TR2011-680.
\section{Credits and Licensing}
\label{sec-5}

@@

  Katana is being written by James Oakley and was designed by Sergey
  Bratus, James Oakley, Ashwin Ramaswamy, Michael Locasto, and Sean
  Smith.
